Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-01-01 survey/survey.tex
continue with Data II

We now have in our hands our desired denotational semantics. We want
to create a programming language, which we call $\Pi$, such that the
types and type combinators map to $\bot, \top, \uplus, \times$, and
such that we have ground terms whose denotation are all $16$ type
isomorphisms of Fig.~\ref{type-isos}. This is completely
straightforward, as we can simply do this literally. To make the
analogy with commutative semirings stand out even more, we will use
$0, 1, +$, and $\times$ at the type level, and will denote
``equivalence'' by $\iso$.  Thus Fig.~\ref{pi-terms} shows the
``constants'' of the language.

@@

\jc{wavefront}

The type-erasure will, naturally, be permutations. But where do the
isomorphisms come from?  Well, inhabitants of types with (0,1,+,*)
form a semiring. So, still inspired by Curry-Howard, look to the
proof-terms of semiring identities -- and voila, a nice set of
isomorphisms! We can name these isomorphisms, AND give them
implementations.

That gives us some ground terms.  Not enough for a PL! So then
we need to be able to 'combine' these. So we add 3 combinators for
3 different kinds of composition.
Gives us a nice, reversible PL for
structured finite types. We call it Pi.

>>>>

We now have in our hands our desired denotational semantics. We want
to create a programming language, which we call $\Pi$, such that the
types and type combinators map to $\bot, \top, \uplus, \times$, and
such that we have ground terms whose denotation are all $16$ type
isomorphisms of Fig.~\ref{type-isos}. This is completely
straightforward, as we can simply do this literally. To make the
analogy with commutative semirings stand out even more, we will use
$0, 1, +$, and $\times$ at the type level, and will denote
``equivalence'' by $\iso$.  Thus Fig.~\ref{pi-terms} shows the
``constants'' of the language. Note how we have continued with
the spirit of Curry-Howard: the terms of $\Pi$ are \emph{proof terms},
but rather than being witnesses of inhabitation, they are witnesses of
equivences. Thus the proof language of commutative semiring equivalences
becomes terms of $\Pi$.

@@

The type-erase of this language \emph{should} be permutations. But
given types $A$ and $B$ that erase to $\fin{m}$ and $\fin{n}$ respectively,
what are $A + B$ and $A * B$ ?  They correspond exactly to
$\fin{m+n}$ and $\fin{m*n}$!  Geometrically, this corresponds to
concatenation for $A + B$, i.e. lining up the elements of $A$ first,
and then those of $B$. For $A * B$, one can picture this as
lining up the elements of $A$ horizontally, those of $B$ vertically
and perpendicular to those of $A$, and filling in the square with
pairs of elements from $A$ and $B$; if one re-numbers these sequentially,
reading row-wise, this gives an enumeration of $\fin{m*n}$.

From here, it is easy to see what, for example, $c_1 \oplus c_2$ must be,
operationally: from a permutation on $\fin{m}$ and another on $\fin{n}$,
create a permutation on $\fin{m+n}$ by having $c_1$ operate on the first
$m$ elements of $A+B$, and $c_2$ operate on the last $n$ elements. 
Similarly, $\swapp$ switches the roles of $A$ and $B$, and thus corresponds
to $\fin{n+m}$. Note how we ``recover'' the commutativity of 
natural number addition from this type isomorphism. Geometrically, $\swapt$
is also rather interesting: it corresponds to matrix transpose!

\jc{it would be nice to have a picture or two here}
